Mapping Phenomena Relevant to Adolescent Emotion Regulation: A Text-mining Systematic Review

### 2021-09-01 presentation_eadp.Rmd
EADP presentation

---
title: "Mapping Phenomena Relevant to Adolescent Emotion Dysregulation: A Text-mining Systematic Review"
author: "Caspar J. Van Lissa<sup>1</sup> (2021), *Adolescent Research Review*, DOI: [10.1007/s40894-021-00160-7](https://link.springer.com/article/10.1007%2Fs40894-021-00160-7)"
date: "<p><sup>1</sup>: Utrecht University, dept. Methodology & Statistics<br>Funded by NWO Veni grant  VI.Veni.191G.090"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
output: 
  ioslides_presentation:
    widescreen: true
    css: slide.css
bibliography      : ["veni_sysrev.bib"]
---

@@

As many as one in five develop psychopathology [@leeAdolescentMentalHealth2014]

@@

Need to know what phenomena are associated with adolescents' emotion regulation

@@

There is a need for the integration of knowledge across research areas and consolidation of empirical research into overarching theory.

@@

According to *Theory Construction Methods*, first step toward a unifying framework is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]

@@

Few tools exist for identifying relevant phenomena (aside from expert opinion)

*This paper argues that systematic reviews are a suitable and relatively more objective method for phenomena detection.*

@@

Narrative reviews are limited by small samples, confirmation bias, and emphasis on positive results [@littellEvidencebasedBiasedQuality2008]

*Text mining systematic review (TMSR):*

* Covers arbitrarily greater corpora
* Derives insights through a transparent, objective, and reproducible process
* Assumes that term frequency indicates relevance of phenomena
* Assumes that co-occurrence frequency indicates putative associations
    + These metrics can be jointly visualized as a network graph
    + Thus "mapping" phenomena relevant to adolescent emotion regulation
    + Co-occurrence graph ~= rudimentary nomological network

## Relevant theory

* Bioecological model [@bronfenbrennerBioecologicalModelHuman2007]
* Transactional model [@sameroffUnifiedTheoryDevelopment2010]
* Hall's notion of "storm and stress"
* Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995]
* Tripartite model [@morrisRoleFamilyContext2007]
* Internalization model of emotional development [@holodynskiDevelopmentEmotionsEmotion2006]
* Polyvagal theory [@porgesOrientingDefensiveWorld1995].
* Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012]
* Process model of emotion regulation [@grossHandbookEmotionRegulation2013]
* Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000]

@@

Phenomena relevant to adolescents' emotion regulation according to theory (a) and narrative reviews (b; transparent nodes indicate constructs also present in the theory).

@@

* Few explicitly address adolescence
* Few have comprehensively addressed important predictors of development in this life stage
* None directly guide contemporary research 
* Theories vary widely in scope: Some are broad and non-specific; others describe a specific phenomenon in detail, but lack a broader perspective.
    + Broad theories can frame any research, specific theories generate hypotheses.
    + It would be beneficial to bridge these levels of analysis.

@@

All code and data available at github.com/cjvanlissa/veni_sysrev

@@

## Screening contd.

Preliminary screening was conducted in Rayyan.

After 367 papers, screening continued in the free open source program ASReview [@van_de_schoot_rens_2020_3828293]

Algorithm: "naive Bayes"

Reverse coding: Excluded articles coded as hits

Screening continued until among the most recently screened 100 papers only 6 were excluded

## Screening contd.

Experienced challenges:

* First version required NVidia CUDA
* Difficult to set up Python environment
* Workaround for corpus with mostly relevant publications
* Would be better to have automatic cutoff

# Analysis 1: Author keywords

## Method

* Keywords extracted by document
* Exclusion filter of methodological terms etc
* The resulting corpus consisted of 5031 documents with 8080 terms
* Dictionary of 108 terms with 464 regular expression queries
* Remaining unique terms: 5292
* Terms exceeding $97.5^{th}$ percentile of negative binomial distribution pruned (= 21+ documents)
* Most pruned terms (4004) occurred only once in the corpus

84 remaining terms, occurring in 4827 documents

## Mapping the literature

* Term co-occurrence matrix, 2498 co-occurrence relationships
* Small coefficients again pruned using $97.5^{th}$ percentile of negative binomial (25+ documents)

106 co-occurrence relationships remained

Terms categorized as cause, outcome, protective factor, or indicator of emotion regulation

# Analysis 2: Abstract text mining

## Method

Nouns and adjectives help derive more interpretable text mining models [@martinMoreEfficientTopic2015]

Nouns capture terms like "emotion"; adjectives capture the "mental" in "mental health"

Used "part-of-speech tagging" (POS-tagging) to identify words' grammatical functions

Stemmed terms to their root

Identified trigrams using `textrank` [@wijffelsTextrankSummarizeText2019]

* E.g., "emotion regulation" is already a bigram

Merged trigrams with original data

## Method contd.

* 5097 documents with 11448 unique words
* Pruned $97.5^{th}$ percentile (6+ documents)
* 850 co-occurrence relationships
* Pruned $97.5^{th}$ percentile (10+ documents)

43 remaining co-occurrence relationships

@@

## Results: Important phenomena

* Emotion regulation and related mental health outcomes among top terms
* Theoretically important themes: *neural*, *parenting*, and *stress*
* Also several under-theorized concepts: *ADHD/CD*, *substance* use, *minority status*
* Substantial correspondence between keywords and abstracts
    + 75% of frequent terms from keywords also present in abstracts
    + 89% of the frequent terms from abstracts present in keywords
* *sex* was more frequent in the Abstracts: sex differences are regularly reported in Abstracts even if they are not mentioned in the keywords
* *parenting* (keywords) was displaced by *mothers* (Abstracts): Parenting is often operationalized as mothering

## Results: Co-occurrence

Keywords:

* Emotion regulation is a central construct to which other constructs are directly linked
* Graph was sparse, with few interconnections between terms (fragmented literature?)

Abstracts:

* Emotion regulation and mental health-related outcomes formed central axis
* Relevant terms connected to this axis, but not directly to emotion regulation
* Only terms directly connected to emotion regulation were *neural*, *mothers*, *attachment* and *ptsd*
* even sparser network

## Discussion

* Both analyses reflected some phenomena from theoretical literature
    + Particularly related to neurodevelopment and socialization
* Mental health-related outcomes feature prominently
    + Emotion regulation implicated in mental health problems [@leeAdolescentMentalHealth2014],
    + This underlines the importance of research in this area

@@

* Developmental disorders
    + E.g., ADHD/CD and autism
* Physical health (sic)
    + sleep, sports, and disability status
* External stressors
    + bullying, stress, PTSD, abuse, violence, life events, historic events (e.g., earthquakes, war), parenting stress, and divorce.
* Structural disadvantage
    + minority status, discrimination, disability status, socio-economic status, adoption status, and sexual diversity
    + Developmental disorders also relevant in this context, as neuroatypical individuals tend to experience social exclusion [@cappadociaBullyingExperiencesChildren2012]

## Undertheorized themes contd.

* Addictive behavior
    + substance use and device use
* Identity and moral development
    + values and personality. 
* Sexual development
    + (biological) sex, puberty, sexual diversity, and romantic relationships

Text mining systematic reviews can thus identify undertheorized themes

## Structural findings

Central axis of emotion regulation and mental health outcomes suggests that these phenomena are closely linked

Sparse networks (fragmented literature)

* Most phenomena directly linked with emotion regulation/mental health
* Absence of terms suggests that they are not well-integrated in the literature
* Sparser network for Abstractslikely reflects greater noise

Substantial correspondence between both networks

* Supports validity of the findings, suggests that automatic keyword extraction identifies relevant constructs

@@

Results reflected familiar phenomena: Suggests validity of the method

Results also revealed under-theorized phenomena: TMSR may complement theory and narrative reviews

Implications for future empirical research:

* Researchers typically rely on theory
* Under-theorized phenomena may be overlooked
* This study offers guidance regarding relevant phenomena to consider as confounders or contributing causes
* Unembedded terms indicate potential knowledge gaps; promising areas of future research

## Theoretical implications

* Sparse graphs indicate fragmented literature
* Important to generate an overarching theory of adolescent emotion regulation
* First step in theory construction methodology is to identify relevant phenomena [@borsboomTheoryConstructionMethodology2020]
* Present study thus serves as starting point for theory development

@@

* This study was more comprehensive than previous narrative reviews [cf. @bariolaChildAdolescentEmotion2011; @coe-odessEmergentEmotionsAdolescence2019]
    + Used systematic search and included much larger corpus
* TMSR can include corpora of arbitrary size, and derive insights by transparant and reproducible procedures

@@

* These text mining techniques do not extract *meaning* from the literature
* Analyses limited to keywords and abstracts
    + Many full text articles are behind paywalls
    + No standardized Application Programming Interface (API)
* Comprehensive access to scientific publications is essential to all meta-scientific research
* Recommend open access publishing and the development of a unified API for article retrieval
* Subjective decisions were made throughout the analysis process
    + creative license is inherent to inductive studies [@wagenmakersCreativityVerificationCyclePsychological2018]
    + There are infinite possible ways to conduct exploratory analyses

@@

Text mining systematic reviews can identify theoretically relevant phenomena,
and map potential associations

* This illustrates validity of the method

Results also revealed undertheorized themes

* E.g., developmental disorders, physical health, external stressors, structural disadvantage, substance use, identity, moral, and
sexual development

Text mining systematic reviews thus complement theory and narrative reviews

>>>>

---
title: "Mapping Phenomena Relevant to Adolescent Emotion Regulation: A Text-mining Systematic Review"
author: "Caspar J. Van Lissa<sup>1</sup> (2021), *Adolescent Research Review*, DOI: [10.1007/s40894-021-00160-7](https://link.springer.com/article/10.1007%2Fs40894-021-00160-7)"
date: "<p><sup>1</sup>: Utrecht University, dept. Methodology & Statistics<br>Funded by NWO Veni grant  VI.Veni.191G.090"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
output: 
  ioslides_presentation:
    widescreen: true
    css: slide.css
bibliography      : ["veni_sysrev.bib"]
---

@@

20% develop psychopathology [@leeAdolescentMentalHealth2014]

@@

What phenomena are associated with adolescents' emotion regulation?

@@

Knowledge across research areas must be integrated and consolidated into overarching theory.

@@

First step in *Theory Construction Methods* is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]

@@

Few tools for identifying relevant phenomena

* aside from expert opinion

*Text mining systematic reviews are a suitable and relatively objective method for phenomena detection.*

@@

**Narrative reviews:** small samples, confirmation bias, emphasis on positive results [@littellEvidencebasedBiasedQuality2008]

**Text mining systematic review (TMSR):**

* No limit on sample size
* Transparent, objective, reproducible process
* Assumes that *term frequency indicates relevance* of phenomena
* Assumes that *co-occurrence frequency indicates association*
    + Co-occurrence graph ~= rudimentary nomological network

## Baseline

Manual selection of key constructs from relevant theory and prior reviews [ @bariolaChildAdolescentEmotion2011; @coe-odessEmergentEmotionsAdolescence2019]:

* Bioecological model [@bronfenbrennerBioecologicalModelHuman2007]
* Transactional model [@sameroffUnifiedTheoryDevelopment2010]
* Hall's notion of "storm and stress"
* Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995]
* Tripartite model [@morrisRoleFamilyContext2007]
* Internalization model [@holodynskiDevelopmentEmotionsEmotion2006]
* Polyvagal theory [@porgesOrientingDefensiveWorld1995].
* Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012]
* Process model of emotion regulation [@grossHandbookEmotionRegulation2013]
* Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000]

@@

(a) Theory (b) narrative reviews; transparent nodes indicate constructs also present in theory.

@@

* Few explicitly address adolescence
* Few comprehensively address predictors of emotion regulation
* None directly guide contemporary research 
* Theories vary widely in scope: Some are broad and non-specific; others describe a specific phenomenon in detail, but lack a broader perspective.
    + Broad theories can frame any research, specific theories generate hypotheses.
    + It would be beneficial to bridge these levels of analysis.

@@

All code and data available at <https://github.com/cjvanlissa/veni_sysrev>

@@

# Method

## Analysis 1: Author keywords

* Extract author keywords
* Exclude methodological terms etc: 5031 documents with 8080 terms
* Dictionary of 108 terms with 464 regular expression queries
* Remaining unique terms: 5292
* Pruned terms exceeding $97.5^{th}$ percentile of negative binomial distribution (= 21+ documents)
* Most pruned terms (4004) occurred only once
* 2498 term co-occurrence relationships
* Pruned using negative binomial (25+ documents)

Final data: **4827 documents** with **84 terms** and **106 co-occurrence relationships**

Categorized causes, outcomes, protective factors, and indicators

## Analysis 2: Abstract text mining

* Selected nouns and adjectives using "part-of-speech tagging"
* Stemmed terms to their root
* Identified trigrams using `textrank` [@wijffelsTextrankSummarizeText2019]
* 5097 documents with 11448 unique words
* Pruned $97.5^{th}$ percentile (6+ documents)
* 850 co-occurrence relationships
* Pruned $97.5^{th}$ percentile (10+ documents)

Final data: **3096 documents** with **71** and **43 co-occurrence relationships**

@@

## Results

* Both analyses reflected some constructs from theoretical literature
    + Particularly related to neurodevelopment and socialization
* Mental health-related outcomes feature prominently
    + Emotion regulation implicated in mental health problems [@leeAdolescentMentalHealth2014],
    + This underlines the importance of research in this area
* Substantial correspondence between keywords and abstracts, suggesting validity of method
* Networks are sparse; few connections among constructs: fragmented literature

@@

* Developmental disorders
* Physical health (sic)
* External stressors
* Structural disadvantage
* Addictive behavior
* Identity and moral development
* Sexual development

@@

* Empirical research relies on theory; under-theorized phenomena may be overlooked
    + This study offers guidance for phenomena to consider as confounders/causes
* Unembedded terms indicate promising areas of future research
* Present study serves as starting point for theory development

@@

* More comprehensive than previous narrative reviews [cf. @bariolaChildAdolescentEmotion2011; @coe-odessEmergentEmotionsAdolescence2019]
    + Used systematic search and included much larger corpus
* TMSR can process samples of any size, derive insights by transparent and reproducible procedures

@@

* These techniques do not extract *meaning* from the literature
* No access to full text articles (paywalls)
* Subjective decisions in analysis process [@wagenmakersCreativityVerificationCyclePsychological2018]

@@

Text mining systematic reviews can complement theory and narrative reviews

* Identify (undertheorized) relevant phenomena
* Map potential associations
